### 2017-07-06 paper/extensions.tex
general edits

As in the case of a.s. termination we start with general mathematical statement 
about LexRSMs. We define a restricted class of LexRSMs with \emph{bounded 
expected conditional increase} property. Recall from 
definition~\ref{def:lexrsm} that 
LexRSM for a stopping time $\stime$ is characterized by the possibility to 
a.s. partition, for each $i\in \Nset_0$, the set $\{\omega\in \Omega\mid 
\stime(\omega)>i \}$ into $d$ sets $L^i_1,\dots,L^i_d$ such that, intuitively, 
on $L^i_j$ the conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is smaller than $\vec{X}_i[j]$, and for all $j'<j$, on $L^i_{j'}$ the 
conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is no larger than $\vec{X}_i[j]$. This leaves the opportunity of conditional 
expectation of $\vec{X}_{i+1}[j]$ being larger than $\vec{X}_i[j]$ on 
$L^i_{j''}$ with $j''$. The conditional expected increase property bounds the 
possibility of this increase.

\begin{definition}
Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be a 
$d$-dimensional LexRSM for some stopping time $\stime$, defined w.r.t. some 
filtration $\{\genfilt_i\}_{i=0}^{\infty}$. We say that 
$\{\vec{X}_{i}\}_{i=0}^{\infty}$ has \emph{$\vec{c}$-bounded expected 
conditional 
increase (ECI),} for 
some non-negative vector $\vec{c}\in\Rset^d$, if for each $i\in \Nset_0 $ and 
each $1\leq 
j \leq d $ it holds 
$\E[\vecseq{X}{i+1}{j}\mid \genfilt_i]\leq \vecseq{X}{i}{j}+\vec{c}[j]$ on 
$L^i_{j''}$, 
for all $j''>j$ 
(here $L^i_1,\dots,L^i_d$ are as in Definition~\ref{def:lexrsm}).
\end{definition}

For LexRSMs with $\vec{c}$-bounded ECI we have the following result. For 
simplicity, 
we formulate the result for 1-LexRSMs, though it is easy to prove analogous 
result for general $\eps$-LexRSMs, $\eps>0$, at the cost of obtaining less 
readable formula.

\begin{theorem}
\label{thm:runtime-bound}
Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be a 
$d$-dimensional LexRSM with $c$-bounded ECI for some stopping time $\stime$. 
Then $\E[\stime]\leq  
\sum_{j=1}^{d}\E[\vecseq{X}{0}{j}]\cdot(\vec{c}[j]+1)^{d-j}$.
\end{theorem}
\begin{proof}
Denote $\noofdecrank_j(\omega)$ the number of steps $i$ in which $\omega\in 
L_j^i$. Since by Theorem~\ref{thm:lexrsm-main} the existence of LexRSM entails 
$\probm(\stime<\infty)=1$, the value $\noofdecrank_j(\omega)$ is a.s. finite. 
We prove that for each $1\leq j \leq d$ it holds $\E[\noofdecrank_j]\leq 
\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\noofdecrank_{j'}]\right) + 
\E[\vecseq{X}{0}{j}].$ Since $\stime(\omega)=\sum_{1\leq j \leq d} 
\noofdecrank_j(\omega)$, for each $\omega\in \Omega$ (and hence, due to 
linearity of expectation $\E[\stime]=\sum_{1\leq j \leq d} 
E[\noofdecrank_j]$), the statement of the 
Theorem follows by an easy induction.

@@

To transfer this mathematical result to probabilistic programs, we want to 
impose a restriction on LexRSM maps that ensures that all components of a 
LexRSM map have, from each reachable configuration, an expected one-step 
increase of at most $c$. Here $c$ can be a constant, but it can also be a value 
that depends on the initial configurations: this is to handle cases where some 
variables are periodically reset to a value related to the initial variable 
values, such as variable $z$ in Figure~\ref{fig:uniint2}. To this end, let 
$\program$ be a \PP{} with a pCFG $\pCFG_\program$ and let 
$\vec{\lem}=(\lem_1,\dots,\lem_d)$ be a 
$d$-dimensional $1$-LexRSM map for $\program$. Consider a $d$-dimensional 
vector 
$\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_d)$ whose each component is an 
expression over variables of 
$\program$. We say 
that $\vec{\lem}$ has 
$\vec{\bar{c}}$-bounded ECI w.r.t. invariant map $\inv$ if the following holds 
for each initial configuration $(\locinit,\vecinit)$ with $\vecinit\in 
\vecinitset$: for 
each 
configuration $(\loc,\vec{x})$ with $\vec{x}\in \inv(\loc)$ and generalized 
transition $\tilde{\tau}$ 
of 
$\pCFG_\program$ outgoing from $(\loc,\vec{x})$ it holds that if $j$ is 
the smallest index such that 
$\tilde{\tau}$ is $1$-ranked by $\lem_j$ from $(\loc,\vec{x})$, then for all 
$j'>j$ the gen. transition $\tilde{\tau}$ is $f$-ranked by $\lem_{j'}$ from 
$(\loc,\vec{x})$, where $f=-c_{j'}(\vecinit)$. From 
Theorem~\ref{thm:runtime-bound} we have the following:

\begin{corollary}
\label{col:runtime-progs}
Let $\program$ be a probabilistic program. Assume that there exists a 
$d$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_d)$ for 
$\program$ supported 
by some 
invariant map $\inv$, scuh that $\vec{\lem}$ has $\vec{\bar{c}}$-bounded ECI 
(w.r.t. $\inv$) 
for some vector of expressions $\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_d)$. 
Then under each scheduler $\sigma$ and for each initial valuation of program 
variables $\vecinit\in\vecinitset$ it holds $\E^{\sigma}_{\vecinit}[\ttime]\leq 
\sum_{j=1}^{d}\lem_j(\locinit,\vecinit)\cdot(\bar{c}_j(\vecinit))^{d-j}$.
\end{corollary}

>>>>

As in the case of a.s. termination we start with general mathematical statement 
about LexRSMs. We define a restricted class of strict LexRSMs with \emph{bounded 
expected conditional increase} property. Recall from 
definition~\ref{def:lexrsm} that strict
LexRSM for a stopping time $\stime$ is characterized by the possibility to 
a.s. partition, for each $i\in \Nset_0$, the set $\{\omega\in \Omega\mid 
\stime(\omega)>i \}$ into $n$ sets $L^i_1,\dots,L^i_d$ such that, intuitively, 
on $L^i_j$ the conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is smaller than $\vec{X}_i[j]$, and for all $j'<j$, on $L^i_{j'}$ the 
conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
is no larger than $\vec{X}_i[j]$. This leaves the opportunity of conditional 
expectation of $\vec{X}_{i+1}[j]$ being larger than $\vec{X}_i[j]$ on 
$L^i_{j''}$ with $j''>j$. The conditional expected increase property bounds the 
possibility of this increase.

\begin{definition}
	\label{def:lexrsm-eci}
Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be an 
$n$-dimensional strict LexRSM for some stopping time $\stime$, defined w.r.t. some 
filtration $\{\genfilt_i\}_{i=0}^{\infty}$. We say that 
$\{\vec{X}_{i}\}_{i=0}^{\infty}$ has \emph{$\vec{c}$-bounded expected 
conditional 
increase (ECI),} for 
some non-negative vector $\vec{c}\in\Rset^d$, if there exists an instance $(\{\vec{X}_{i=0}^{\infty},\{L_1^i,\dots,L_{n+1}^i\}_{i=0}^{\infty})$ of the strict LexRSM (i.e. $L^i_{n+1}=\emptyset$ for all $i$) such that for each $i\in \Nset_0 $ and 
each $1\leq 
j \leq n $ it holds 
$\E[\vecseq{X}{i+1}{j}\mid \genfilt_i]\leq \vecseq{X}{i}{j}+\vec{c}[j]$ on 
$L^i_{j''}$, 
for all $j''>j$ 
(here $L^i_1,\dots,L^i_n$ are as in Definition~\ref{def:lexrsm}).
\end{definition}

For strict LexRSMs with $\vec{c}$-bounded ECI we have the following result. For 
simplicity, 
we formulate the result for 1-LexRSMs, though it is easy to prove analogous 
result for general $\eps$-LexRSMs, $\eps>0$, at the cost of obtaining less 
readable formula.

\begin{theorem}
\label{thm:runtime-bound}
Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be an 
$n$-dimensional strict LexRSM with $c$-bounded ECI for some stopping time $\stime$. 
Then $\E[\stime]\leq  
\sum_{j=1}^{n}\E[\vecseq{X}{0}{j}]\cdot(\vec{c}[j]+1)^{n-j}$.
\end{theorem}
\begin{proof}
Fix an instance $(\{\vec{X}_{i=0}^{\infty},\{L_1^i,\dots,L_{n+1}^i\}_{i=0}^{\infty})$ satisfying Definition~\ref{def:lexrsm-eci}. Denote $\noofdecrank_j(\omega)$ the number of steps $i$ in which $\omega\in 
L_j^i$. Since by Theorem~\ref{thm:lexrsm-main} the existence of strict LexRSM entails 
$\probm(\stime<\infty)=1$, the value $\noofdecrank_j(\omega)$ is a.s. finite for all $1\leq j \leq n$. 
We prove that for each $1\leq j \leq n$ it holds $\E[\noofdecrank_j]\leq 
\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\noofdecrank_{j'}]\right) + 
\E[\vecseq{X}{0}{j}].$ Since $\stime(\omega)=\sum_{1\leq j \leq n} 
\noofdecrank_j(\omega)$, for each $\omega\in \Omega$ (and hence, due to 
linearity of expectation $\E[\stime]=\sum_{1\leq j \leq n} 
E[\noofdecrank_j]$), the statement of the 
Theorem follows by an easy induction.

@@

To transfer this mathematical result to probabilistic programs, we want to 
impose a restriction on LexRSM maps that ensures that all components of a 
LexRSM map have, from each reachable configuration, an expected one-step 
increase of at most $c$. Here $c$ can be a constant, but it can also be a value 
that depends on the initial configurations: this is to handle cases where some 
variables are periodically reset to a value related to the initial variable 
values, such as variable $z$ in Figure~\ref{fig:uniint2}. To this end, let 
$\program$ be a \PP{} with a pCFG $\pCFG_\program$ and let 
$\vec{\lem}=(\lem_1,\dots,\lem_n)$ be an 
$n$-dimensional $1$-LexRSM map for $\program$. Consider an $n$-dimensional 
vector 
$\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_n)$ whose each component is an 
expression over variables of 
$\program$. We say 
that $\vec{\lem}$ has 
$\vec{\bar{c}}$-bounded ECI w.r.t. invariant map $\inv$ if the following holds 
for each initial configuration $(\locinit,\vecinit)$ with $\vecinit\in 
\vecinitset$: for 
each 
configuration $(\loc,\vec{x})$ with $\vec{x}\in \inv(\loc)$ and generalized 
transition $\tilde{\tau}$ 
of 
$\pCFG_\program$ outgoing from $(\loc,\vec{x})$ it holds that if $j$ is 
the smallest index such that 
$\tilde{\tau}$ is $1$-ranked by $\lem_j$ from $(\loc,\vec{x})$, then for all 
$j'>j$ the gen. transition $\tilde{\tau}$ is $f$-ranked by $\lem_{j'}$ from 
$(\loc,\vec{x})$, where $f=-c_{j'}(\vecinit)$. From 
Theorem~\ref{thm:runtime-bound} we have the following:

\begin{corollary}
\label{col:runtime-progs}
Let $\program$ be a probabilistic program. Assume that there exists an 
$n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_d)$ for 
$\program$ supported 
by some 
invariant map $\inv$, scuh that $\vec{\lem}$ has $\vec{\bar{c}}$-bounded ECI 
(w.r.t. $\inv$) 
for some vector of expressions $\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_n)$. 
Then under each scheduler $\sigma$ and for each initial valuation of program 
variables $\vecinit\in\vecinitset$ it holds $\E^{\sigma}_{\vecinit}[\ttime]\leq 
\sum_{j=1}^{n}\lem_j(\locinit,\vecinit)\cdot(\bar{c}_j(\vecinit))^{n-j}$.
\end{corollary}

### 2017-07-06 paper/lexicographic-programs.tex
minor changes
finished sec on expected runtime
re-shuffled the proof of the main theorem a bit
general edits

\section{Applying Lexicographic Supermartingales to Probabilistic Programs}

@@

\textbf{[PETR: DEFINE LOCATION TERMINATION]}

@@

\begin{theorem}
\label{thm:lexrsm-programs}
Let $\program$ be a probabilistic program. Assume that there exists an $\eps>0$ 
and an $n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for 
$\program$ supported 
by some 
invariant map $\inv$. 
Then $\program$ terminates almost surely.
\end{theorem}
\begin{proof}
Let $\sigma$ be any measurable scheduler and $\vecinit\in\vecinitset$ any 
initial variable valuation in $\program$.
We define an $n$-dimensional stochastic process 
$\{\vec{X}_{i}\}_{i=0}^{\infty} $ on the probability space 
$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ such 
that for each 
$i\geq 0$ and $1\leq j 
\leq n$ and each run $\run$ we put $\vecseq{X}{i}{j}(\run) = 
\lem_j(\cfg{\sigma}{i}(\run))$. We claim that $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is an $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
the process is real-valued, componentwise non-negative, and adapted to the 
canonical filtration of $\natfilt$. It remains to prove that condition (3) in 
Definition~\ref{def:lexrsm} is satisfied. To this end, for each $i\geq 0$ we 
define an almost-sure partition of the set $\{\run\in \OmegaRun\mid 
\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting $L^i_j$ to be 
the set of all runs $\run$ such that $\ttime(\run)>i$ and for $\run$ the index 
$j$ is the smallest one such that the $(i+1)$-th transition on $\run$ is ranked 
by $\lem_j$ from $\cfg{\sigma}{i}(\run)$. Due to 
definition of an $\eps$-LexRSM map such a $j$ exists for all 
$\run\in\{\ttime>i\}$ and hence we indeed have a partition. It remains to prove 
that irrespective of the initial choice of $\sigma$ and $\vecinit$ it holds, 
for each $1\leq j 
\leq n$ and $j'<j$, that $\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq X_i[j]-\eps $ on $L_j^i$ and 
$\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq X_i[j] $ on $L_{j'}^i$. This follows easily from 
the definition of $L^{i}_1,\dots,L^{i}_n$ and from the definition of a 
transition being $\eps$-ranked by $\lem_j$.

Since  $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is an $\eps$-LexRSM for $\ttime$, from Theorem~\ref{thm:lexrsm-programs} it 
follows that $\probm^{\sigma}_{\vecinit}\ttime <\infty) =1$, irrespective of 
$\sigma$ and $\vecinit$.
\end{proof}

@@

\begin{example}
Consider the program in Figure~\ref{fig:inftime}. It is easy to see that it 
terminates a.s., but the expected termination time is infinite: to see this, 
note that that the expected value of variable $x$ upon reaching the second loop 
is $\frac{1}{2}\cdot 1 + \frac{1}{4}\cdot 2 + \frac{1}{8}\cdot 4 + \cdots = 
\frac{1}{2}+\frac{1}{2}+\frac{1}{2}+\cdots=\infty$ and that the time needed to 
get out of the second loop is equal to the value of $x$ upon entering the loop. 
However, a.s. termination of the program is proved by a 2-dimensional LinLexRSM 
$\lem$ pictured in the figure.
\end{example}

>>>>

\section{Applying Lexicographic Supermartingales to Probabilistic Programs}
\label{sec:lex-programs}

@@

%\textbf{[PETR: DEFINE LOCATION TERMINATION]}

@@

\begin{theorem}
\label{thm:lexrsm-programs}
Let $\program$ be a probabilistic program. Assume that there exists an $\eps>0$ 
and an $n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for 
$\program$ supported 
by some 
invariant map $\inv$. 
Then $\program$ terminates almost surely.
\end{theorem}
\begin{proof}
Let $\sigma$ be any measurable scheduler and $\vecinit\in\vecinitset$ any 
initial variable valuation in $\program$.
We define an $n$-dimensional stochastic process 
$\{\vec{X}_{i}\}_{i=0}^{\infty} $ on the probability space 
$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ such 
that for each 
$i\geq 0$ and $1\leq j 
\leq n$ and each run $\run$ we put $\vecseq{X}{i}{j}(\run) = 
\lem_j(\cfg{\sigma}{i}(\run))$. We claim that $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is a strict $n$-dimensional $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
the process is real-valued, componentwise non-negative, and adapted to the 
canonical filtration of $\natfilt$. It remains to prove that condition (3) in 
Definition~\ref{def:lexrsm} is satisfied. To this end, for each $i\geq 0$ we 
define an almost-sure partition of the set $\{\run\in \OmegaRun\mid 
\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting $L^i_j$ to be 
the set of all runs $\run$ such that $\ttime(\run)>i$ and for $\run$ the index 
$j$ is the smallest one such that the $(i+1)$-th transition on $\run$ is ranked 
by $\lem_j$ from $\cfg{\sigma}{i}(\run)$. Due to 
definition of an $\eps$-LexRSM map such a $j$ exists for all 
$\run\in\{\ttime>i\}$ and hence we indeed have a partition (so $L^i_{n+1}=\emptyset$ for all $i$). It remains to prove 
that irrespective of the initial choice of $\sigma$ and $\vecinit$ it holds, 
for each $1\leq j 
\leq n$ and $j'<j$, that $\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq X_i[j]-\eps $ on $L_j^i$ and 
$\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
\natfilt_i]\leq X_i[j] $ on $L_{j'}^i$. This follows easily from 
the definition of $L^{i}_1,\dots,L^{i}_n$ and from the definition of a 
transition being $\eps$-ranked by $\lem_j$.

Since  $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
is an $\eps$-LexRSM for $\ttime$, from Theorem~\ref{thm:lexrsm-programs} it 
follows that $\probm^{\sigma}_{\vecinit}(\ttime <\infty) =1$, irrespective of 
$\sigma$ and $\vecinit$.
\end{proof}

@@

\begin{example}
\label{ex:infinite-time}
Consider the program in Figure~\ref{fig:inftime}. It is easy to see that it 
terminates a.s., but the expected termination time is infinite: to see this, 
note that that the expected value of variable $x$ upon reaching the second loop 
is $\frac{1}{2}\cdot 1 + \frac{1}{4}\cdot 2 + \frac{1}{8}\cdot 4 + \cdots = 
\frac{1}{2}+\frac{1}{2}+\frac{1}{2}+\cdots=\infty$ and that the time needed to 
get out of the second loop is equal to the value of $x$ upon entering the loop. 
However, a.s. termination of the program is proved by a 2-dimensional LinLexRSM 
$\lem$ pictured in the figure.
\end{example}
